Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)

### 2017-05-10 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Added markdown to select covariates based on weighted random effect based modeling with BIC criteria

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)

@@

thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'

@@

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
#"RibozeroBatch", "FlowcellBatch",
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E.NA = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts.NA = PROCESSED_COUNTS$filteredExprMatrix$counts
PROCESSED_COUNTS$filteredExprMatrix$counts.NA[log.mat] = NA
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$E.NA
CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$E.zero))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E.zero, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E.NA) %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.zero)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E.zero, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. 
1. "Dx.Tissue" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
2. Adding Instituion and Gender a priori to variable selection as they are correlated with Dx.Tissue
```{r iterative.normalisation, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender');

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'Individual_ID'))
covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects and Individual_ID is chosen as random effect'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
  correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

  # Re-calculate voom weights with correlation of random effects
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                              design=DM1$design, plot=F,
                              block = COVARIATES$Individual_ID, 
                              correlation = correlation$cor)

  # Fit linear model using new weights and new design
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
  ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = list()
  RESIDUAL.GENE_EXPRESSION$E.NA = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E.NA)
  RESIDUAL.GENE_EXPRESSION$E.zero = RESIDUAL.GENE_EXPRESSION$E.NA
  RESIDUAL.GENE_EXPRESSION$E.zero[is.na(RESIDUAL.GENE_EXPRESSION$E.NA)] = 0

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'Individual_ID'))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  tmp.expr = RESIDUAL.GENE_EXPRESSION$E.zero
  tmp.expr = tmp.expr[apply(tmp.expr,1,var) !=0,]
  residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
                                                                          c(postAdjustCovars, 'Individual_ID'))]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
tmp.expr = RESIDUAL.GENE_EXPRESSION$E.zero
tmp.expr = tmp.expr[apply(tmp.expr,1,var) !=0,]
residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
                                              COVARIATES, 
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION$E.NA = RESIDUAL.GENE_EXPRESSION$E.NA + 
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION$E.zero))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION$E.zero, scale.=T, center = T)

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION$E.zero)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Post adjusted design matrix
DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
correlation <- parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$Individual_ID, 
                    correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
FIT = lmFit(VOOM.WEIGHTS)

# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC"),
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

  normalizationType	= 'CQN'
)

# Code
CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

@@

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts.NA %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$E.NA %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

>>>>

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)')
```

@@

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)

@@

thisFileName <- 'CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd'

@@

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 5 cpm counts in at least 100% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx.Tissue), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID %>% droplevels()],
                                                   MIN_GENE_CPM=5, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=1)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t

# Back calculate counts for further analysis
LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
```

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
NEW.COUNTS[log.mat] = NA
CQN.GENE_EXPRESSION$E[log.mat] = NA
```

### Co-expression distribution
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
cr = cor(t(tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
PC <- prcomp(tmp, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
tree = hclust(as.dist(t(tmp)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp, include = F}
dev.off()
gc()
```

### Distribution of samples (log cpm)
```{r lcpm.dist}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E) %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
p
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
preAdjustedSigCovars = runPCAandPlotCorrelations(tmp, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){

  # Get design matrix
  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]

  # Estimate voom weights
  cnt[is.na(cnt)] = 0
  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights

  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
                              .combine = rbind,
                              .packages = c('lme4', 'dplyr'),
                              .export = c('voomWeights')) %dopar% {

                                # Perform weighted mixed linear models for every gene
                                expr = data.frame(Gx = lcpm[i,])
                                data = cbind(expr,
                                             designMat[rownames(expr),],
                                             Individual_ID = groupingVar)
                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }

# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){

  varsSelected = c()
  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
  while(length(covarsToSelect) != 0){
    allModelBIC = foreach(i = 1:length(covarsToSelect),
                          .combine = c,
                          .packages = pkgs,
                          .export = c('fitMixedEffectModel')) %dopar% {

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           covars[colnames(cnt),'Individual_ID'], 
                                                           lcpm, 
                                                           cnt, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

                            return(list(modelBIC))
                          }
    allModelBIC = plyr::join_all(allModelBIC)

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = better - worse)

    kable(nChange)

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > round(dim(cnt)[1]*0.05)) %>%
      dplyr::top_n(1, nchange)

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('Dx.Tissue', 'Institution', 'Gender')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase1.covars = c('PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase2.covars = c('LibraryBatch', 'RibozeroBatch', 'FlowcellBatch')
vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected",
                  "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", "PercentAligned")
vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`

### Normalisation
1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect
```{r iterative.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate voom weights
tmp = NEW.COUNTS
tmp[is.na(tmp)] = 0
VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)

# Estimate consensus correlation coefficient
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(tmp, 
                            design=DM1$design, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
tmp = RESIDUAL.GENE_EXPRESSION
tmp[is.na(tmp)] = 0
residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

### Coexpression of residual expression 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
tmp = RESIDUAL.GENE_EXPRESSION
tmp[is.na(tmp)] = 0
cr = cor(t(tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(tmp, scale.=T, center = T)

@@

tree = hclust(as.dist(t(tmp)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

  normalizationType	= 'CQNandMS'
)

# Code
CODE <- Folder(name = "ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN (Model Selection)", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

@@

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
